---
title: "Lobster-ECOL README"
format: gfm
execute: 
  echo: false
  warning: false
  message: false
  fig.height: 1
  fig.width: 3
  comment: ""
  fig.align: center
---

```{r}
library(here)
library(sf)
library(rnaturalearth)
library(tidyverse)
library(gmRi)
library(showtext)

# Read Regions in
proj_path <- cs_path("mills", "Projects/Lobster ECOL")
vts_poly    <- read_sf(str_c(proj_path, "Ecological Data/Spatial_Boundaries/VTSsurvey_nearshore_area.geojson"))
res_shapes  <- cs_path("res", "Shapefiles")
epu_path    <- str_c(res_shapes, "EPU/")
gom_poly    <- read_sf(str_c(epu_path, "individual_epus/GOM.geojson"))
gb_poly     <- read_sf(str_c(epu_path, "individual_epus/GB.geojson"))

# Statebounds
canada <- ne_states("Canada", returnclass = "sf") %>% filter(name_en %in% c("Nova Scotia", "Quebec", "New Brunswick", "Prince Edward Island"))
ne_us <- ne_states("United States of America", returnclass = "sf") %>% filter(name %in% c("Maine", "Massachusetts", "Vermont", "New Hampshire", "Connecticut", "New York", "Rhode Island"))

# Map theme
theme_set(theme_bw() + 
            map_theme(
              plot.title = element_text(family = "Avenir", face = "bold", size = 14),
              axis.text = element_text(family = "Avenir", size = 11),
              plot.text = element_text(family = "Avenir")))
```




```{r}
#| label: fonts-config

# Path to the directory containing the font file (replace with your actual path)
font_dir <- paste0(system.file("stylesheets", package = "gmRi"), "/GMRI_fonts/Avenir/")

# Register the font
font_add(
  family = "Avenir",
  file.path(font_dir, "LTe50342.ttf"),
  bold = file.path(font_dir, "LTe50340.ttf"),
  italic = file.path(font_dir, "LTe50343.ttf"),
  bolditalic = file.path(font_dir, "LTe50347.ttf"))

# Load the font
showtext::showtext_auto()

```

# Lobster-ECOL

This is the ecological data processing code for the Lobster ecology project. This project's scope ranges across spatial scales working outward from a nearshore area that is sampled with Maine's ventless trap survey (3nm), to an area that is further offshore but state managed (6nm), and ultimately to t shelf-scale area that is sampled by the Federal Government's fisheries independent surveys.

At each of these spatial scales physical and ecological metrics have processed for further use in research.Code in this repository covers the acquisition and data processing of physical and ecological datasets at the following scales and from the following sources:

# A.) Nearshore Scale Maine + New Hampshire Metrics:

```{r}
#| echo: false

ggplot() +
  geom_sf(data = vts_poly, fill = gmri_cols("blue"), alpha = 0.4) +
  geom_sf(data = ne_us) +
  geom_sf(data = canada) +
  coord_sf(xlim = c(-71.5, -66), ylim = c(42.5, 45)) +
  labs(title = "Nearshore Scale")

```

### Local Currents

 1. Maine Coastal Current, Turnoff
 
> Source: FVCOM

#### Sea Surface Temperature

 1. SST Anomalies
 2. Days over 20 C
 3. Days within 12-18 C

 > Source: FVCOM

#### Bottom Temperature

 1. BT Anomalies
 2. Days over 20 C
 3. Days within 12-18 C

> Source: FVCOM


---

# B.) State/Regional Scale: Gulf of Maine, SNE & GB Metrics:

```{r}
ggplot() +
  geom_sf(data = gom_poly, fill = gmri_cols("orange"), alpha = 0.4) +
  geom_sf(data = gb_poly, fill = gmri_cols("gmri green"), alpha = 0.4) +
  geom_sf(data = ne_us) +
  geom_sf(data = canada) +
  coord_sf(xlim = c(-72, -65), ylim = c(40, 45)) +
  labs(title = "State/Regional Scales")
```


#### Sea Surface Temperature
 1. SST Anomalies   
 2. Days over 20 C
 3. Days within 12-18 C
 
> Source: FVCOM

#### Bottom Temperature
 1. BT Anomalies
 2. Days over 20 C
 3. Days within 12-18 C
 
 > Source: FVCOM

#### Primary Productivity
 1. Annual PAR and photosynthetic efficiency 

> Source: Ecodata

#### Zooplankton
 1. Annual abundance by taxa

> Source: Ecodata

#### Lobster Predator Indices
 1. Predator abundance
 2. Predator size spectra

> Sources: NEFSC & ME/NH Survey

#### Predator exploitation rate
 1. Based on rate as calculated in the 2020 lobster stock assessment as the annual catch of lobster divided by the estimate of population abundance

 > Source: ASMFC 2020 lobster stock assesment

---


# C.) Northeast US Shelf-wide Scale:



#### Gulf Stream Position
 > Source: Ecodata


#### Sea Surface Temperature
 1. SST Anomalies
 2. Days over 20 C
 3. Days within 12-18 C

> Source: FVCOM

#### Bottom Temperature
 1. BT Anomalies
 2. Days over 20 C
 3. Days within 12-18 C 
 
> Source: FVCOM

#### Salinity

> Only if requested
> Source: FVCOM

#### Currents
1. Relative inflow of GS vs. Scotian Shelf water at NE Channel 

> Source: ecodata::slopewater


# Lobster-ECOL Quarto Docs Directory

[Area Subsetting in R with {FVCOM}](https://adamkemberling.github.io/Lobster-ECOL/R/FVCOM_prep/FVCOM_Area_Subsetting_Demo.html)

[Exploring Monthly File Contents and Structures](https://adamkemberling.github.io/Lobster-ECOL/R/FVCOM_prep/FVCOM_GMRI-Monthly-Inventory-Check.html)

[Maine Coastal Current Processing](https://adamkemberling.github.io/Lobster-ECOL/R/FVCOM_prep/FVCOM_MaineCoastalCurrent.html)