---
title: "Lobster-ECOL README"
format: gfm
execute: 
  echo: false
  warning: false
  message: false
  fig.height: 2
  fig.width: 4
  comment: ""
  fig.align: center
---

```{r}
library(here)
library(sf)
library(rnaturalearth)
library(tidyverse)
library(gmRi)
library(showtext)

# Read Regions in
proj_path <- cs_path("mills", "Projects/Lobster ECOL")

# Load Shapefiles for inshore/offshore
poly_paths <- cs_path("mills", "Projects/Lobster ECOL/Spatial_Defs")

inshore_areas <- map(
  setNames(
    list.files(str_c(poly_paths, "inshore_areas"), full.names = T),
    str_remove(list.files(str_c(poly_paths, "inshore_areas")), ".geojson")),
  function(x){read_sf(x)}
)

offshore_areas <- map(
  setNames(
    list.files(str_c(poly_paths, "offshore_areas"), full.names = T),
    str_remove(list.files(str_c(poly_paths, "offshore_areas")), ".geojson")),
  function(x){read_sf(x)}
)

# Statebounds
canada <- ne_states("Canada", returnclass = "sf")# %>% filter(name_en %in% c("Nova Scotia", "Quebec", "New Brunswick", "Prince Edward Island"))
ne_us <- ne_states("united states of america", returnclass = "sf") %>% 
  filter(postal %in% c("VT", "ME", "RI", "MA", "CT", "NH", "NY", "MD", "VA", "NJ", "DE", "NC", "PA", "WV"))

# Map theme
theme_set(theme_bw() + 
            map_theme(
              plot.title = element_text(family = "Avenir", face = "bold", size = 14),
              axis.text = element_text(family = "Avenir", size = 11),
              plot.text = element_text(family = "Avenir")))
```




```{r}
#| label: fonts-config

# Path to the directory containing the font file (replace with your actual path)
font_dir <- paste0(system.file("stylesheets", package = "gmRi"), "/GMRI_fonts/Avenir/")

# Register the font
font_add(
  family = "Avenir",
  file.path(font_dir, "LTe50342.ttf"),
  bold = file.path(font_dir, "LTe50340.ttf"),
  italic = file.path(font_dir, "LTe50343.ttf"),
  bolditalic = file.path(font_dir, "LTe50347.ttf"))

# Load the font
showtext::showtext_auto()
theme_set(theme_bw() + map_theme(text = element_text(family = "Avenir")))

```

# Lobster-ECOL

This is the ecological data processing code for the Lobster ecology project. This project's scope ranges across spatial scales working outward from a nearshore area that is sampled with Maine's ventless trap survey (3nm), to an area that is further offshore but state managed (6nm), and ultimately to the shelf-scale area that is sampled by the Federal Government's fisheries independent surveys.

At each of these spatial scales physical and ecological metrics have processed for further use in research. Code in this repository covers the acquisition and data processing of physical and ecological datasets at the following scales and from the following sources:


```{r}
#| label: environmental indicators table
#| eval: true

tribble(
  ~"Area",     ~"Metric",                        ~"Frequency",   ~"Data Source", 
  "Nearshore", "Sea Surface Temperature",         "Daily",            "FVCOM",
  "Nearshore", "Sea Bottom Temperature",          "Daily",            "FVCOM",
  "Nearshore", "Sea Surface Salinity",            "Daily",            "FVCOM",
  "Nearshore", "Sea Bottom Salinity",             "Daily",            "FVCOM",
  "Nearshore", "Maine Coastal Current Continuity","Daily",            "FVCOM",
  "Nearshore", "Days within 12-18C",              "Daily",            "FVCOM",
  "Nearshore", "Days over 20C",                   "Daily",            "FVCOM",
  "Offshore", "Sea Surface Temperature",         "Daily",            "FVCOM",
  "Offshore", "Sea Bottom Temperature",          "Daily",            "FVCOM",
  "Offshore", "Sea Surface Salinity",            "Daily",            "FVCOM",
  "Offshore", "Sea Bottom Salinity",             "Daily",            "FVCOM",
  "Offshore", "Days within 12-18C",              "Daily",            "FVCOM",
  "Offshore", "Days over 20C",                   "Daily",            "FVCOM",
  "Offshore", "Primary Productivity (ppd, ppd ratio)", "Seasonal",         "ecodata::chl_pp",
  "Offshore", "Primary Productivity (Chl-A)",            "Seasonal",         "ecodata::chl_pp",
  "Offshore", "Zooplankton Community Index",     "Annual",  "ecodata::zoo_abundance_anom",
  "Offshore", "Lobster Predator Abundance",      "Seasonal",         "Trawl Surveys",
  "Offshore", "Lobster Predator Size Spectra",   "Seasonal",         "Trawl Surveys",
  "Shelf-Scale", "Gulf Stream Position Index",   "Annual/Monthly","ecodata::gsi",
  "Shelf-Scale", "Northeast Channel Water Mass", "Annual",           "ecodata::slopewater") %>% 
  #dplyr::select(-Area) %>% 
  knitr::kable() #%>% 
  # kableExtra::pack_rows(group_label = "Inshore Areas: 0-3 Nautical miles", 1, 7) %>% 
  # kableExtra::pack_rows(group_label = "Offshore Regions: 12+ Nautical miles", 8, 18) %>% 
  # kableExtra::pack_rows(group_label = "Shelf-Scale Processes", 19, 20) 




# Slopewater ecodata::slopewater
# GSI ecodata::gsi & ecodata::gsi_old
# Calanus Stage? ecodata::plot_calanus_stage
# Zooplankton: ecodata::zoo_diversity, ecodata::zoo_abundance_anom, ecodata::zoo_regime
# Primary productivity ecodata::chl_pp

```





# A.) Nearshore Scale


```{r}
#| label: fig-nearshore-scale-map
#| echo: false

ggplot() +
  geom_sf(data = bind_rows(inshore_areas), aes(fill = area_id), alpha = 0.4) +
  geom_sf(data = ne_us) +
  geom_sf(data = canada) +
  scale_fill_gmri() +
  coord_sf(xlim = c(-78, -66), ylim = c(35.5, 45)) +
  labs(title = "Nearshore Scale: < 12nm", fill = "")

```



---

# B.) Offshore & Regional Scale

 1. Gulf of Maine & Georges Bank
 2. Southern New England

```{r}
#| label: fig-offshore-scale-map

ggplot() +
  geom_sf(data = bind_rows(offshore_areas), aes(fill = Region), alpha = 0.4) +
  geom_sf(data = ne_us) +
  geom_sf(data = canada) +
  scale_fill_gmri() +
  coord_sf(xlim = c(-78, -66), ylim = c(35.5, 45)) +
  labs(title = "Offshore Scale: > 12nm", fill = "")
```


---


# C.) Northeast US Shelf-wide Scale

```{r}

ggplot() +
  geom_sf(data = st_union(bind_rows(offshore_areas)), aes(fill = "Shelf-Scale"), alpha = 0.4) +
  geom_sf(data = ne_us) +
  geom_sf(data = canada) +
  scale_fill_gmri() +
  coord_sf(xlim = c(-78, -66), ylim = c(35.5, 45)) +
  labs(title = "Offshore Scale: > 12nm", fill = "")
```



# Lobster-ECOL Quarto Docs Directory

[Area Subsetting in R with {FVCOM}](https://adamkemberling.github.io/Lobster-ECOL/R/FVCOM_prep/fvcom_general_methods/FVCOM_Area_Subsetting_Demo.html)

[Exploring Monthly File Contents and Structures](https://adamkemberling.github.io/Lobster-ECOL/R/FVCOM_prep/gmri_inventory_exploration/FVCOM-GMRI-Monthly-Inventory-Check.html)

[Interpolating Point Values](https://adamkemberling.github.io/Lobster-ECOL/R/FVCOM_prep/Survey_Locations_FVCOMTemp_Interpolation.html)

[Timeseries for Areas of Interest](https://adamkemberling.github.io/Lobster-ECOL/R/FVCOM_prep/FVCOM_Regional_Temperature_Timeseries.html)

[Maine Coastal Current Processing](https://adamkemberling.github.io/Lobster-ECOL/R/FVCOM_prep/MaineCoastalCurrent/FVCOM_MaineCoastalCurrent_PCA.html)

[FVCOM Temperature Regime Shifts with STARS](https://adamkemberling.github.io/Lobster-ECOL/R/regime_tests/STARS_FVCOM.html)

