---
title: "Seasonal Mean Interpolations"
description: | 
  Linear Interpolation of Seasonal Mean Surface + Bottom Temperatures for Trawl+VTS Survey Locations
date: "Updated on: `r Sys.Date()`"
format: 
  html:
    code-fold: true
    code-tools: true
    df-print: kable
    self-contained: true
execute: =
  echo: true
  warning: false
  message: false
  fig.align: "center"
  comment: ""
---


```{r}

# Packages
{
library(raster)        # netcdf data as raster
library(sf)            # vector spatial mapping/operations
library(fvcom)         # fvcom mesh and variable extractions
library(ncdf4)         # netcdf support
library(tidyverse)     # data wrangling and plotting
library(gmRi)          # color schemes and cloud storage paths
library(patchwork)     # plot arrangement
library(rnaturalearth) # coastlines and state polygons
library(geometry)      # bathycentric coordinates
library(Matrix)        # matrix algebra
library(sysfonts)      # font support
}


# Paths + conflicts
conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
# proj_path <- cs_path("mills", "Projects/Lobster ECOL")
# Path to project folder
proj_path <- cs_path("mills", "Projects/Lobster ECOL/lobster_data")

# # Path(s) to state trawl survey data
state_surveys <- str_c(proj_path, "state_trawl/")


source(here::here("R/FVCOM_Support.R"))
theme_set(theme_bw() + map_theme())

# Shapefiles
new_england <- ne_states("united states of america", returnclass = "sf") %>% 
  filter(postal %in% c("VT", "ME", "RI", "MA", "CT", "NH", "NY", "MD", "VA", "NJ", "DE", "NC", "PA", "WV"))
canada <- ne_states("canada", returnclass = "sf")


# Gom EPU to clip
res_shapes <- cs_path("res", "Shapefiles")
epu_path <- str_c(res_shapes, "EPU/")
gom_poly <- read_sf(str_c(epu_path, "individual_epus/GOM.geojson"))

```




```{r}
#| label: style-sheet
#| results: asis

# Use GMRI style
use_gmri_style_rmd()

```



```{r}
#| label: fonts-config
#| echo: false

# Path to the directory containing the font file (replace with your actual path)
font_dir <- paste0(system.file("stylesheets", package = "gmRi"), "/GMRI_fonts/Avenir/")

# Register the font
font_add(
  family = "Avenir",
  file.path(font_dir, "LTe50342.ttf"),
  bold = file.path(font_dir, "LTe50340.ttf"),
  italic = file.path(font_dir, "LTe50343.ttf"),
  bolditalic = file.path(font_dir, "LTe50347.ttf"))

# Load the font
showtext::showtext_auto()

```

# About: Seasonal FVCOM for Survey Data

This quarto steps through the data prep for attaching the appropriate seasonal mean surface and bottom conditions using FVCOM data from the nearest three mesh nodes.

Survey coordinates from both state and federal trawl and trap surveys will be prepared this way, with sampling seasons adjusted for the survey seasons.


## Seasonal Mean Temperature and Salinity for State+Federal Lobster Survey Locations

Before Claire left, I wrote down these notes:
 - Resolve the join issue that was leading to survey locations with missing FVCOM data (could be a data filtering thing during survdat prep, or a lat/lon rounding thing)
 - For the trawl locations we want the average conditions over 3-month periods, not the day-of conditions
 - For VTS survey data this will be the July & August 2-month average
 
 
 
 
### Load Surveys


For loading in the station info, we want to standardize the columns and keep with us the bare minimum we need to join back to the catch data.

This should include the survey program, some unique tow ID, season, date, latitude, and longitude. For cases where "ID" is not an explicit column already, it will be constructed to create a unique identifier. The unique identifier will be used to join the FVCOM data back to the survey data. This should perform better than coordinate fields which may mis-match along the way if there are is any rounding that occurs.

Starting latitude and longitude will be used for the positioning. These will be renamed to `lon` & `lat` for laziness.


Lastly a new column `program` will be created to identify the sampling program that the data came from.

#### 1. Federal trawl survey

For the federal trawl survey data, the unique ID column will be a concatenation of `svvessel`,`cruise6`, `station`, `stratum`, & `tow`. 
 
```{r}
#| label: load federal survey data

### Load NEFSC Trawl Data

# Load the trawl data
fed_trawl_path <- cs_path("res", "nmfs_trawl/SURVDAT_current")
# tidy it a little - drops inshore strata
fed_trawl_dat <- read_rds(str_c(fed_trawl_path, "survdat_lw.rds")) %>% 
  pluck("survdat") %>% 
  janitor::clean_names()

# Get distinct time/date/tow/locations
fed_trawl_stations <- fed_trawl_dat %>% 
  mutate(
    station_id = str_c(svvessel, cruise6, station, stratum, tow, sep = "-"),
    date = as.Date(est_towdate),
    program = "NMFS Trawl Survey"
  ) %>% 
  distinct(program, station_id, season, date, lat, lon)

gt::gt(head(fed_trawl_stations))
```
 
#### 2. Maine & New Hampshire Trawl Survey

For Maine and New Hampshire the `id` column will be a concatenation of `survey`, `year`, `region`, `depth_stratum`, & `tow_number`.
 

```{r}
#| label: menh trawl data

# Maine NH Trawl Survey - ID
menh_path <- str_c(state_surveys, "ME_trawldata/")
menh_trawl <- read_csv(str_c(menh_path, "MaineDMR_Trawl_Survey_Tow_Data_2024-05-17.csv"))  %>% 
  janitor::clean_names()
menh_trawl_stations <- menh_trawl  %>% 
  mutate(
   station_id = str_c(survey, year, region, depth_stratum, tow_number),
   date = as.Date(start_date),
   program = "MENH Trawl Survey") %>% 
  distinct(program, station_id, season, date, lat = start_latitude, lon = start_longitude)

gt::gt(head(menh_trawl_stations))
```


#### 3. Massachusetts Trawl Survey

The `id` for the MA state trawl survey will be a concatenation of `vessel`, `cruise6`, `stratum`, & `station`.


```{r}
#| label: mass trawl data

# Mass Trawl
mass_path  <- cs_path("res", "MA_Trawl/Pull_20240716/Manipulated")
mass_trawl <- read_csv(str_c(mass_path, "MADMF_SVSTA_SW_2024.csv"))  %>% 
  janitor::clean_names()
mass_trawl_stations <- mass_trawl %>% 
  mutate(
    date = as.Date(
      str_c(
        year, 
        str_pad(month, side = "left", pad = "0", width = 2), 
        str_pad(day, side = "left", pad = "0", width = 2),
        sep = "-")),
    station_id = str_c(vessel, cruise6, station, sep = "-"),
    program = "MA Trawl Survey") %>% 
  distinct(program, station_id, season, date, lat = start_lat, lon = start_lon)

gt::gt(head(mass_trawl_stations))
```

#### 4. Rhode Island Trawl

Rhode Island state survey has a `trawl_identity` field. Some of these stations lack lat/lon info so there will be fewer records than this identifier.

RI's new `station_id` will use `season` and `trawl_identity` to make sure the field reads as a string anywhere in the future.


```{r}
#| label: RI trawl Data

# Rhode Island
load(str_c(state_surveys, "RI_LOBSTER_TRAWL_012324.Rdata"))
# Loads RI_Stations & RI_Lengths
ri_trawl <- RI_Stations %>% 
  janitor:: clean_names() 

ri_trawl_stations <- ri_trawl %>% 
  drop_na(lat, lon) %>% 
  mutate(
    date = as.Date(
      str_c(year, 
        str_pad(month, side = "left", pad = "0", width = 2), 
        str_pad(day, side = "left", pad = "0", width = 2),
      sep = "-")),
    station_id = str_c(season, trawl_identity, sep = "-"),
    program = "RI Trawl Survey") %>% 
  distinct(program, station_id, season, date, lat, lon)

gt::gt(head(ri_trawl_stations))

```

For the NEAMAP data, there is an ID column already, so we can use that. We can make a date column from ymd info, and then pull the coordinates.

```{r}
#| label: Neamap trawl data

# NEAMAP
load(str_c(state_surveys, "NEAMAP Trawl.Rdata"))

# Loads these two objects
# glimpse(Stations) # Details on where stations happened
# glimpse(Lengths)  # Details on the catch

# Tidy it up
neamap_stations <- Stations %>% 
  janitor::clean_names() %>% 
  mutate(
    station_id = id,
    date = as.Date(
    str_c(
      year, 
      str_pad(month, side = "left", pad = "0", width = 2), 
      str_pad(day, side = "left", pad = "0", width = 2),
      sep = "-")),
    program = "NEAMAP Survey") %>% 
  distinct(program, station_id, season, date, lon = longitude, lat = latitude)

gt::gt(head(neamap_stations))

```


For the New Jersey data the start and end coordinates need to be changed to something. I will first see if NAD1983-UTM18 is a good projected coordinate system, that or the state plane crs.



```{r}
#| label: NJ trawl data

# New Jersey
nj_trawl <- readxl::read_xlsx(str_c(state_surveys, "NJ AM LOBSTER TRAWL SURVEY_RAW DATA_2023 TERMINAL.xlsx")) %>% 
  janitor::clean_names() 

# Pull Stations
nj_stations <- nj_trawl %>% 
  mutate(
    station_id = str_c(vessel, id, sep = "-"),
    yrmoda = as.character(yrmoda),
    date = as.Date(yrmoda, format = "%Y%m%d"),
    program = "NJ Trawl Survey") %>% #glimpse()
  distinct(program, station_id, date, lat = slat, lon = slong) %>% 
  drop_na(lon, lat)

# Change coordinates to degrees
# UTM Zone 18N — Covers New Jersey and the surrounding offshore waters.
# EPSG: 32618 (WGS84) or 26918 (NAD83).
# New Jersey State Plane (NAD83)
# EPSG: 3424 (US Feet) or 6527 (Meters)
nj_stations %>% 
  # st_as_sf(coords = c("lon", "lat"), crs = st_crs(32618), remove = F) %>% # Nope
  # st_as_sf(coords = c("lon", "lat"), crs = st_crs(26918), remove = F) %>% # Nope
  # st_as_sf(coords = c("lon", "lat"), crs = st_crs(6527), remove = F) %>% # Nope
  st_as_sf(coords = c("lon", "lat"), crs = st_crs(3424), remove = F) %>% # Nope
  ggplot() + 
  geom_sf() +
  geom_sf(data = new_england) +
  coord_sf(xlim = c(-78, -65), ylim = c(35, 42), crs = 4326)


# When are the seasons?
nj_stations %>% 
  mutate(month = month.abb[lubridate::month(date)],
         month = factor(month, levels = month.abb)) %>% 
  count(month)
```



```{r}
#| label: load VTS points

# Path to resources
vts_path <- cs_path("mills", "Projects/Lobster/VTS_fromASMFC")

# Maine
load(str_c(vts_path, "VTS_data.Rdata"))

# Mass
load(str_c(vts_path, "VTS_MA_Proc_240201 all data for standardized index.Rdata"))

# Trips_MA # Date/time details for trawl
# Trawls_MA # Trawl details
# Sites_MA # Site Locations
# Traps_MA # trap details ventless/vented
# Lobsters_MA # Individual lobsters from survey


# Need to join Trips (for date) and Trawls for the location details

# # We have cases where the same trip is given a different stock
# Trawls %>% filter(TripId == Trips[319,]$TripId)
# Trips %>% filter(TripId == Trawls[25969,]$TripId)

# There are six cases where a trawlID has different lat/lon values across different records

vts_trips <- bind_rows(
  list(
    "MA VTS Survey" = inner_join(
      x = Trips_MA, y = Trawls_MA, 
      by = join_by(TripId)) %>% 
      mutate(Fisher = as.character(Fisher)),
    "ME VTS Survey" = inner_join(
      Trips, Trawls, 
      join_by(TripId), 
      relationship = "many-to-many")),
  .id = "program") %>% 
  distinct(program, TrawlId, date = Date, lon = Longitude, lat = Latitude) %>% 
  drop_na(lon, lat)
 
# # Can join them this way too, weird
# vts_trips <- full_join(
#     x = bind_rows(Trips, mutate(Trips_MA, Fisher = as.character(Fisher))),
#     y = bind_rows(Trawls, Trawls_MA), 
#     by = join_by(TripId), 
#   relationship = "many-to-many") %>% 
#   mutate(program = "Ventless Trap Survey") %>% 
#   distinct(program, TrawlId, SiteId, date = Date, lon = Longitude, lat = Latitude) 



# # Mass Lobster Stations?
# load(str_c(state_surveys, "MassDMF_Lobster_StationsNLengths_1979_2023_230718.Rdata"))
# Stations # year, season, coords, ID
# Lengths # ID, sex, length, NumCal

```



```{r}
# Plot them all
all_trawl <- bind_rows(
  list(mass_trawl_stations, menh_trawl_stations, neamap_stations, fed_trawl_stations, ri_trawl_stations)
)


ggplot() + 
  geom_point(
    data = all_trawl, aes(lon, lat, color = program), 
    shape = 3, size = 0.5, alpha = 0.2) +
  geom_point(
    data = vts_trips, 
    aes(lon, lat, color = "Ventless Trap Survey"), 
    shape = 3, size = 0.5, alpha = 0.2) +
  geom_sf(data = new_england) +
  coord_sf(xlim = c(-75.75, -67), ylim = c(35, 44.75)) +
  guides(color = guide_legend(override.aes = list(alpha = 1, shape = 15, size = 2))) +
  #scale_color_gmri() +
  facet_wrap(~program) +
  scale_color_gmri() +
  #rcartocolor::scale_color_carto_d(palette = "Prism") +
  labs(x = "Longitude", y = "Latitude", color = "Survey Program")



```


### Load FVCOM Inventory

```{r}
# Surface and Bottom only, from Dr. Li
fvcom_path <- cs_path("res", "FVCOM/Lobster-ECOL")


# Here are the files we have, loop over them later
fvcom_surfbot_files <- setNames(
  list.files(fvcom_path, full.names = T, pattern = ".nc"),
  str_remove(list.files(fvcom_path, full.names = F, pattern = ".nc"), ".nc"))


# Test File: GOM3 1978
# Load some daily FVCOM that we downloaded and averaged
fvcom_yrx <- nc_open(fvcom_surfbot_files["gom3_1978"])

# Get the mesh itself as a simple feature collection
gom3_mesh <- get_mesh_geometry(fvcom_yrx, what = 'lonlat')


```


### Check Point-Mesh Coverage Overlap

The following map shows the coverage overlap between trawl survey locations and FVCOM GOM3.

```{r}

# Map everything
ggplot() +
  geom_sf(data = gom3_mesh, 
          alpha = 0.2, linewidth = 0.05, color = "gray30") +
  geom_sf(
    data = st_as_sf(all_trawl, coords = c("lon", "lat"), crs = 4326),
    aes(color = "Trawl Surveys"),
    shape = 3, size = 0.2, alpha = 0.2) +
  geom_sf(
    data = st_as_sf(vts_trips, coords = c("lon", "lat"), crs = 4326),
    aes(color = "VTS Surveys"),
    shape = 3, size = 0.2, alpha = 0.2) +
  geom_sf(data = new_england) +
  geom_sf(data = canada) +
  theme(legend.position = "right") +
  scale_fill_gmri() +
  coord_sf(xlim = c(-78, -58), ylim = c(34, 46)) +
  theme_bw() + map_theme() +
  labs(
    title = "Coverage overlap of FVCOM and Sample Points",
    fill = "Area")


```


# Matching FVCOM Nodes + Times to Station Coordinates


To pull FVCOM surface and bottom conditions we first need to identify the indexing in space for the neighboring points, and then the start+end indexes in time for each season. This will need to be done for all `r nrow(bind_rows(all_trawl, vts_trips))` locations.

The following code approaches identifying the correct date indices to loop/map through this interpolation process.



## Get the Proper Time Index for Each File


```{r}
#| label: surfbot date matching

# Put the survey programs together
all_stations <- bind_rows(all_trawl, vts_trips) %>% 
  mutate(
    season = tolower(season),
    season = if_else(str_detect(program, "VTS"), "summer", season),
    season = case_match(
      season,
      "s" ~ "spring",
      "spring" ~ "spring",
      "f" ~ "fall",
      "fall" ~ "fall",
      "summer" ~ "summer",
      .default = season
  ))


# Get the dates within each file
fvcom_dates <- map_dfr(
  fvcom_surfbot_files,
   function(x){
     x_fvcom <- nc_open(x)
     timez <- ncvar_get(x_fvcom, "Times")
     nc_close(x_fvcom)
     return(
       data.frame("fvcom_date" = timez) %>% 
         mutate(time_idx = row_number()))
   }, .id = "fv_file")



# Join them by date to look at their matches
station_date_matches <- all_stations %>% 
  left_join(
    mutate(fvcom_dates, fvcom_date = as.Date(fvcom_date)),
    join_by(date == fvcom_date))
```







## Get Proper Element and Node Weights for Points


```{r}
#| label: triangulate points function

# Function to add the linear interpolation weights based on node coordinates
triangulate_and_weight <- function(pts_sf, fvcom_mesh){

    # Identify the triangles that overlap each point:
    # Use st_join to assign elem, p1, p2, p3 IDs to the points
    pts_assigned <- st_join(
      st_transform(pts_sf, st_crs(fvcom_mesh)),
      gom3_mesh, 
      join = st_within) %>% 
      drop_na(elem)
  

    # Iterate over the rows to add weights:
    pts_weighted <- pts_assigned %>% 
     base::split(., seq_len(nrow(.))) %>%
     purrr::map_dfr(function(pt_assigned){
    
      # Subset the relevant triangle from st_join info
      triangle_match <- fvcom_mesh[pt_assigned$elem,]
      
      # Build matrices for point to interpolate & of surrounding points:
    
      # Matrix for triangle
      # Use the triangles node coordinates from the sf geometries
      # Creates 3x3 matrix: 
      # row1 x coords, 
      # row 2, y coords, 
      # row three rep(1,3)
      
      node_vertices <- t(
        st_coordinates(triangle_match[1,])[1:3,1:3]
        )
      
      # Make matrix from the points:
      # creates 3x1 matrix: x, y, 1
      point_coords <- matrix(
        c(st_coordinates(pt_assigned[1,]), 1), 
        nrow = 3)
      
      #### For Linear Interpolation:
      
      # Get inverse of the matrix
      inverse_coordmat <- solve(node_vertices)
      
      # Solve for the weights
      node_wts <- inverse_coordmat %*% point_coords %>%
        t() %>% 
        as.data.frame() %>% 
        setNames(c("p1_wt", "p2_wt", "p3_wt"))
      
      # Return with dataframe
      bind_cols(pt_assigned, node_wts)
    
    
    })
    # End Rowwise
    return(pts_weighted)
}


```



```{r}
#| label: get weights for each point

# Make the trawl locations an sf dataframe
stations_sf <- station_date_matches %>% 
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = F)


# Run for all points:
stations_weighted <- triangulate_and_weight(
  pts_sf = stations_sf, 
  fvcom_mesh = gom3_mesh) %>% 
  st_drop_geometry()
```
 
 
# Interpolations at Proper Time Step(s)

For seasonal averages we need to pull a range of dates and then average over that period of time.


 
### Determine the "when" for Seasonal Averages

Most trawl surveys are performed twice a year in the "spring" and "fall". For the federal survey spring begins in March and is mostly done by April, with fall ocurring primarily in Sep-Oct.

For the state surveys "spring" is more centered on May, and fall centered on September or October. This varies by program.

 The VTS survey is a summertime program, with sampling happening fairly evenly across June-August, and then to a lesser extent in Sept. & Oct.
 

```{r} 

# Fall   = Sep-Nov
# Spring = Mar-May
# Both are sampled more heavily in first two months, then as needed in the third
# ex. March and April have 4x more stations than May

all_stations %>% 
  mutate(doy = lubridate::yday(date),
         flat_date = as.Date("2000-01-01") + doy-1) %>% 
  ggplot(aes(flat_date, program)) +
  geom_point(
    shape = "l", size = 2, alpha = 0.25, color = "gray30",
    position = position_jitter(height = 0.2)) +
  labs(x = "Date")


```
 

For each survey we want the average condition over the period of time when these surveys are taking place. To be consistent across the different surveys we will be simplifying seasons to be consistent across surveys:

Spring = March-May
Fall   = September-November
Summer = June-August


The following table will be used as a key for each survey program's seasonal coverage, and provide the starting index (as day of year) to start and finish reading data along the time dimension.

 
```{r}
#| label: setup seasonal indexes

# Set up Season Indexes
season_indexing <- tribble(
  ~"season", ~"start",      ~"end",
  "spring",  "2000-03-01",  "2000-05-30",
  "summer",  "2000-06-01",  "2000-08-30",
  "fall",    "2000-09-01",  "2000-11-30") %>% 
  mutate(
    start = as.Date(start),
    start = lubridate::yday(start),
    end = lubridate::yday(end))

```



## Datewise Interpolation
 
This step can be looped on each date/year to minimize the amount of fvcom file opening/closing. Within each year we need to identify which timestep to extract data at, and then iterate on them.


```{r}
#' Interpolate Values from FVCOM Mesh at Timestep
#' 
#' @description Takes a dataframe row containing node and time indices and interpolation weights and returns that contains time index from which to interpolate.
#' 
#' Should contain the following columns:
#' time_idx = integer timestep to use for interpolation
#' p1 = integer index value for node 1 surrounding the interpolation location
#' p2 = integer index value for node 2 surrounding the interpolation location
#' p3 = integer index value for node 3 surrounding the interpolation location
#' p1_wt = linear interpolation weight for p1
#' p2_wt = linear interpolation weight for p1
#' p3_wt = linear interpolation weight for p1
#'
#' @param interp_details dataframe row containing time index, node index, and node weight information columns
#' @param fvcom_nc FVCOM netcdf file to extract values from
#' @param fvcom_varid = String indicating variable in FVCOM netcdf to interpolate values with
#' @param var_out String to use as variable name in returned dataframe
#'
#' @return
#' @export
#'
#' @examples
# interpolate_seasonal_timestep <- function(
#     interp_details, 
#     fvcom_nc, 
#     fvcom_varid, 
#     var_out, 
#     season_indexing){
#         
#     # Get the values of the variable of interest as vector
#     node_vals <- ncvar_get(
#       nc = fvcom_nc, 
#       varid = fvcom_varid, 
#       start = c(1, 1),
#       count = c(-1, -1))
#     #return(dim(node_vals))
#     
#     # Make a dataframe with the dates, and the three nodes, multiplied by their weights
#     p1 <- node_vals[interp_details[["p1"]], ] * interp_details[["p1_wt"]]
#     p2 <- node_vals[interp_details[["p2"]], ] * interp_details[["p2_wt"]]
#     p3 <- node_vals[interp_details[["p3"]], ] * interp_details[["p3_wt"]]
#     interp_timeseries <- data.frame(
#       "date" = ncvar_get(nc = fvcom_nc, varid = "Times")) %>% 
#       mutate({{var_out}} := p1 + p2 + p3)
#     
#     
#     # Apply optional time-period integration
#     if(missing(season_indexing)){
#       return(interp_timeseries)
#       } else if (is.null(season_indexing)) {
#       message("Argument 'season_indexing' was supplied as NULL.")
#     } else {
#         # This works:
#         seasonal_means <-  pmap_dfr(
#           .l = season_indexing,
#           .f = function(season, start, end) {
#               data.frame("season" = season) %>%
#               mutate(
#                 {{var_out}} := mean(interp_timeseries[c(start:end), var_out])
#           )
#       })
#       
#       return(seasonal_means)
#     }
#   
# }

```



### Test Out One Location Season

```{r}
# test_nc <- nc_open(fvcom_surfbot_files[[1]])
# 
# # Get all dates in the file:
# test_daily <- interpolate_seasonal_timestep(
#   interp_details = stations_weighted[1,], 
#   fvcom_nc = test_nc, 
#   fvcom_varid = "surface_t", 
#   var_out = "surf_temp")
# 
# 
# 
# # Return an everaage over some seasonal period
# # Average over season
# test_seasons <- interpolate_seasonal_timestep(
#   interp_details = stations_weighted[1,], 
#   fvcom_nc = test_nc, 
#   fvcom_varid = "surface_t", 
#   var_out = "surf_temp", 
#   season_indexing = season_indexing) %>% 
#   mutate(date = case_match(
#     season, 
#     "spring" ~ as.Date("1978-04-15"),
#     "summer" ~ as.Date("1978-07-15"),
#     "fall" ~ as.Date("1978-10-15")
#   ))
# 
# 
# ggplot() +
#   geom_line(data = test_daily, aes(as.Date(date), surf_temp), alpha = 0.2) +
#   geom_point(data = test_seasons, aes(as.Date(date), surf_temp, color = season)) + 
#   labs(title = "Seasonal Averaging Test")


```




## Iterate on Years to Interpolate All Points

Now that we can pull/interpolate for a specific timestep, we can loop over the yearly files and obtain surface and bottom temperatures. I'm using yearly timesteps to loop because we have yearly files, this way we only need to open each year once, then slice out values at the corresponding timesteps within them.

```{r}
# Combine surface and bottom extraction into one function, 
# will be so much faster

# interpolate_seasonal_surfbot <- function(
#     interp_details, 
#     fvcom_nc, 
#     fvcom_varid1, 
#     var1_out, 
#     fvcom_varid2, 
#     var2_out, 
#     season_indexing){
#         
#     # Get the values of the variables of interest
#     var1_vals <- ncvar_get(
#       nc = fvcom_nc, 
#       varid = fvcom_varid1, 
#       start = c(1, 1),
#       count = c(-1, -1))
#     var2_vals <- ncvar_get(
#       nc = fvcom_nc, 
#       varid = fvcom_varid2, 
#       start = c(1, 1),
#       count = c(-1, -1))
#     #return(dim(node_vals))
#     
#     # Make a dataframe with the dates, and the three nodes, multiplied by their weights
#     p1_idx <- interp_details[["p1"]]
#     p2_idx <- interp_details[["p2"]]
#     p3_idx <- interp_details[["p3"]]
#     p1_wt  <- interp_details[["p1_wt"]]
#     p2_wt  <- interp_details[["p2_wt"]]
#     p3_wt  <- interp_details[["p3_wt"]]
#     # Pull var1 for the three adjacent nodes, weight them
#     var1_p1 <- var1_vals[p1_idx, ] * p1_wt
#     var1_p2 <- var1_vals[p2_idx, ] * p2_wt
#     var1_p3 <- var1_vals[p3_idx, ] * p3_wt
#     # Pull var2 for the three adjacent nodes, weight them
#     var2_p1 <- var2_vals[p1_idx, ] * p1_wt
#     var2_p2 <- var2_vals[p2_idx, ] * p2_wt
#     var2_p3 <- var2_vals[p3_idx, ] * p3_wt
#     # Assemble as one dataframe
#     interp_timeseries <- data.frame(
#       "date" = ncvar_get(nc = fvcom_nc, varid = "Times")) %>% 
#       mutate(
#         {{var1_out}} := var1_p1 + var1_p2 + var1_p3,
#         {{var2_out}} := var2_p1 + var2_p2 + var2_p3)
#     
#     
#     # Apply optional time-period integration
#     if(missing(season_indexing)){
#       return(interp_timeseries)
#       } else if (is.null(season_indexing)) {
#       message("Argument 'season_indexing' was supplied as NULL.")
#     } else {
#         # This works:
#         seasonal_means <-  pmap_dfr(
#           .l = season_indexing,
#           .f = function(season, start, end) {
#               data.frame("season" = season) %>%
#               mutate(
#                 {{var1_out}} := mean(interp_timeseries[c(start:end), var1_out]),
#                 {{var2_out}} := mean(interp_timeseries[c(start:end), var2_out])
#           )
#       })
#       
#       return(seasonal_means)
#     }
#   
# }




# # Test
# interpolate_seasonal_surfbot(
#   interp_details = stations_weighted[1,], 
#   fvcom_nc = test_nc, 
#   fvcom_varid1 = "surface_t", 
#   fvcom_varid2 = "bottom_t", 
#   var1_out = "surf_temp", 
#   var2_out = "bot_temp", 
#   season_indexing = season_indexing)
```

```{r}
# # Operate over years
# trawl_fvcom_temps <- stations_weighted %>% 
#   drop_na(time_idx) %>% 
#   mutate(year = year(date)) %>% 
#   #filter(year == 2000) %>% 
#   split(.$year) %>% 
#   # Loop over years of data
#   map_dfr(
#     
#     .f = function(samples_year_x){
#     
#       # Pull the file name details
#       nc_name <- samples_year_x[["fv_file"]][[1]][[1]]
#       # Open the corresponding Netcdf
#       fvcom_yr_x <- nc_open(fvcom_surfbot_files[[nc_name]])
#     
#     
#     # Just do every location
#     # We need this:
#     # the seasonal average, which we can filter to just use the season we care about
#     seasonal_interpolations <- pmap_dfr(
#       .l = samples_year_x,
#       .f = function(station_id, p1, p2, p3, p1_wt, p2_wt, p3_wt, season, ...) {
#         
#         # Get the surface temperature
#         surv_season <- season
#         seasonal_means <- interpolate_seasonal_surfbot(
#             interp_details = data.frame(
#               "p1" = p1,
#               "p2" = p2,
#               "p3" = p3,
#               "p1_wt" = p1_wt,
#               "p2_wt" = p2_wt,
#               "p3_wt" = p3_wt), 
#             fvcom_nc = fvcom_yr_x, 
#             fvcom_varid1 = "surface_t", 
#             fvcom_varid2 = "bottom_t", 
#             var1_out = "surf_temp", 
#             var2_out = "bot_temp", 
#             season_indexing = dplyr::filter(
#               season_indexing, season == surv_season)) %>% 
#           mutate(station_id = station_id, .before = "season")
#       })
#         
#         
#     return(seasonal_interpolations)
#     
#    
#     
#   })
```


```{r}
# # Did it work?, should be same number of records
# trawl_fvcom_temps %>% nrow()
# stations_weighted %>% nrow()
# stations_weighted %>% drop_na(time_idx) %>% nrow()
```

# Can We Speed it Up: Yes

The above function takes much longer than necessary because we're pulling the data for all nodes three times

Make a new function that will take the output from ncvar_get, instead of opening it for all the points.

The first function: `subset_nc_var` that takes our indexing and quickly subsets the relevant data. This takes optional additional indexing information for indexing on multiple dimensions

```{r}

#' @title Extract Netcdf Data with Index List
#' 
#' @description Subsets data from the results of ncvar_get(). Accepts single or
#' multiple indexes for one or more dimensions using a list. List order should
#' match the relevant dimensions for the variable.
#' 
#' If no indexing is supplied for a relevant dimension, all elements are returned.
#'
#' @param nc_array 
#' @param index_list 
#'
#' @returns
#' @export
#'
#' @examples
subset_nc_var <- function(nc_array, index_list) {
  
  # Get the number of dimensions
  dims <- length(dim(nc_array))
  
  # Create a list of indices, defaulting to ":" (keeping all elements)
  full_index_list <- rep(list(quote(expr = )), dims)  
  
  # Update with user-specified indices
  for (dim_idx in seq_along(index_list)) {
    if (!is.null(index_list[[dim_idx]])) {
      full_index_list[[dim_idx]] <- index_list[[dim_idx]]
    }
  }
  
  # Use do.call to apply indexing dynamically
  extracted_data <- do.call(`[`, c(list(nc_array), full_index_list, list(drop = TRUE)))
  
  return(extracted_data)
}



# This will get done outside
# Get the values of the variables of interest
test_var_vals <- ncvar_get(
    nc = test_nc, 
    varid = "surface_t", 
    start = c(1, 1),
    count = c(-1, -1))


# Just one index
subset_nc_var(test_var_vals, index_list = list(5))

# Two indices, one node and a range of times
subset_nc_var(test_var_vals, index_list = list(5, 10:20))

```


```{r}
#| label: general interpolation function


# Making the dynamic dimensions work
# Maybe add the st_intersection here? - na
interpolate_seasonal <- function(
    interp_details, 
    fvcom_var_array, 
    fvcom_times,
    var_out, 
    season_indexing){
     
    # Pull the node and weight information for the point location 
    # where values are being interpolated
    node1_idx <- interp_details[["p1"]]
    node2_idx <- interp_details[["p2"]]
    node3_idx <- interp_details[["p3"]]
    node1_wt  <- interp_details[["p1_wt"]]
    node2_wt  <- interp_details[["p2_wt"]]
    node3_wt  <- interp_details[["p3_wt"]]
    
    
    
    # Pull var1 for the three adjacent nodes, weight them
    node1_vec <- subset_nc_var(fvcom_var_array, index_list = list(node1_idx)) * node1_wt
    node2_vec <- subset_nc_var(fvcom_var_array, index_list = list(node2_idx)) * node2_wt
    node3_vec <- subset_nc_var(fvcom_var_array, index_list = list(node3_idx)) * node3_wt
  
    # Assemble as one dataframe
    interp_timeseries <- data.frame(
      "date" = fvcom_times) %>% 
      mutate({{var_out}} := node1_vec + node2_vec + node3_vec)
    
    
    # Apply optional time-period integration
    if(missing(season_indexing)){
      return(interp_timeseries)
      } else if (is.null(season_indexing)) {
        message("Argument 'season_indexing' was supplied as NULL.")
    
        } else {
          # Run the average for the season(s) in season index
          seasonal_means <-  pmap_dfr(
            .l = season_indexing,
            .f = function(season, start, end) {
                data.frame("season" = season) %>%
                mutate(
                  {{var_out}} := mean(interp_timeseries[c(start:end), var_out], na.rm = T))
      
            })
      
      return(seasonal_means)
    }
  
}


# Test this version
interpolate_seasonal(
  interp_details = stations_weighted[2,], 
  fvcom_var_array = test_var_vals, 
  fvcom_times = ncvar_get(test_nc, "Times"), 
  var_out = "surf_temp", 
  season_indexing = season_indexing)
```





```{r}
# Full new flow:
Sys.time()

# Operate over years
trawl_fvcom_temps_new <- stations_weighted %>% 
  drop_na(time_idx) %>% 
  mutate(year = year(date)) %>% 
  #filter(year == 2000) %>% 
  split(.$year) %>% 
  
  # Loop over years of data
  map_dfr(
    
    .f = function(samples_year_x){
    
      # Pull the file name details
      nc_name <- samples_year_x[["fv_file"]][[1]][[1]]
      
      # Open the corresponding Netcdf
      fvcom_yr_x <- nc_open(fvcom_surfbot_files[[nc_name]])
      
      # Pull Surface
      surf_temp_yrx <- ncvar_get(
        nc = fvcom_yr_x, 
        varid = "surface_t", 
        start = c(1, 1),
        count = c(-1, -1))
      
      # Pull Bottom
      bot_temp_yrx <- ncvar_get(
        nc = fvcom_yr_x, 
        varid = "bottom_t", 
        start = c(1, 1),
        count = c(-1, -1))
      
      # Pull Dates
      yrx_dates <- ncvar_get(fvcom_yr_x, "Times")
      
      # Close thje netcdf and return the data
      nc_close(fvcom_yr_x)
    
      
      # For each location, get relevant seasonal average
      seasonal_interpolations <- pmap_dfr(
        .l = samples_year_x,
        .f = function(station_id, p1, p2, p3, p1_wt, p2_wt, p3_wt, season, ...) {
          
          
          # Test the new version:
          surv_season <- season
          relevant_season <-  season_indexing %>% 
            filter(season == surv_season)
         
          # Get the surface temperature
           seasonal_stemp_means <- interpolate_seasonal(
            interp_details = data.frame(
                "p1" = p1,
                "p2" = p2,
                "p3" = p3,
                "p1_wt" = p1_wt,
                "p2_wt" = p2_wt,
                "p3_wt" = p3_wt), 
            fvcom_var_array = surf_temp_yrx, 
            fvcom_times = yrx_dates, 
            var_out = "surf_temp", 
            season_indexing = relevant_season) %>% 
            mutate(station_id = station_id, .before = "season")
          
           # Get Bottom Temps
           seasonal_btemp_means <- interpolate_seasonal(
            interp_details = data.frame(
                "p1" = p1,
                "p2" = p2,
                "p3" = p3,
                "p1_wt" = p1_wt,
                "p2_wt" = p2_wt,
                "p3_wt" = p3_wt), 
            fvcom_var_array = bot_temp_yrx, 
            fvcom_times = yrx_dates, 
            var_out = "bot_temp", 
            season_indexing = relevant_season)
           
           # Combine
           bind_cols(seasonal_stemp_means, select(seasonal_btemp_means, bot_temp))
          
        })
          
      # Return the seasonal data
      return(seasonal_interpolations)
    
   
    
  })

Sys.time()
```




```{r}
# Did it work?, should be same number of records
trawl_fvcom_temps_new 
stations_weighted %>% drop_na(time_idx) %>% nrow()


```


## Export



```{r}
# # # Save VTS Bottom temperatures

# save to project path
fvcom_processed_path <- cs_path("mills", "Projects/Lobster ECOL/FVCOM_processed/point_location_temperatures")

write_csv(
  x = trawl_fvcom_temps_new,
  file = str_c(fvcom_processed_path, "survey_locations_seasonal_surfbottemps.csv"))

```



### Diagnostics - Missing Locations

```{r}

```

