---
title: Gulf of Maine Warming Closeread
date: "Updated on: `r Sys.Date()`"
format: closeread-html
execute:
  echo: false
  warning: false
  message: false
  comment: ""
---



```{r}

# Packages
{
library(sf)            # vector spatial mapping/operations
library(tidyverse)     # data wrangling and plotting
library(gmRi)          # color schemes and cloud storage paths
library(patchwork)     # plot arrangement
library(rnaturalearth) # coastlines and state polygons
library(sysfonts)      # font support
library(ggtext)
library(scales)
}


# Paths + conflicts
conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")

# Shapefiles
new_england <- ne_states("united states of america", returnclass = "sf") %>% 
  filter(postal %in% c("VT", "ME", "RI", "MA", "CT", "NH", "NY", "MD", "VA", "NJ", "DE", "NC", "PA", "WV"))
canada <- ne_states("canada", returnclass = "sf")

# File paths for various extents based on "apershing_gulf_of_maine"
region_paths <- get_timeseries_paths(
  region_group = "gmri_sst_focal_areas", 
  box_location = "cloudstorage")


deg_f <- "\u00b0F"
deg_c <- "\u00b0C"
```




```{r}
#| label: style-sheet
#| results: asis

# Use GMRI style
use_gmri_style_rmd()

```



```{r}
#| label: fonts-config
#| echo: false

# Path to the directory containing the font file (replace with your actual path)
font_dir <- paste0(system.file("stylesheets", package = "gmRi"), "/GMRI_fonts/Avenir/")

# Register the font
font_add(
  family = "Avenir",
  file.path(font_dir, "LTe50342.ttf"),
  bold = file.path(font_dir, "LTe50340.ttf"),
  italic = file.path(font_dir, "LTe50343.ttf"),
  bolditalic = file.path(font_dir, "LTe50347.ttf"))

# Load the font
showtext::showtext_auto()

theme_set(theme_gmri(base_family = "Avenir", title_family = "Avenir"))

```


# Gulf of Maine Warming with Closeread

I've been working on visualizations around this topic for a number of years, but I wanted to try something different to walk through my developing understanding of regional dynamics.

This document uses the [quarto closeread extension](https://closeread.dev). 

`Closeread`is a custom [Quarto](https://quarto.org) extension which can be used to create a storyscrolling experience for html documents.

This is my first attempt at making one. This exercise has two/three goals:

1. A show-and-tell of a new tool
2. An attempt to evaluate whether this is worth the time investment
3. Show some Ocean current or Lobster-ECOL regime results on a map


# Gulf of Maine Warming Dynamics


:::{.cr-section}



In 2015, a study from Dr. Andrew Pershing and colleagues highlighted the rapid warming of the Gulf of Maine and its possible effects on cod populations.[@cr-areamap]{scale-by="1.25"} 

> Over the past decade, sea surface temperatures in the Gulf of Maine increased faster than 99% of the global ocean. - Pershing et al. 2015 



:::{#cr-areamap}


```{r}
# Bathymetric Contour Data
bathy <- raster::raster(str_c(cs_path("res","Bathy/"), "ETOPO1/NEShelf_Etopo1_bathy.tiff")) 

# Make a terra copyfor ggplot using tidyterra
elev_terra <- terra::rast(bathy)

# Polygon Path
poly_path <- region_paths[["apershing_gulf_of_maine"]][["shape_path"]]
region_extent <- read_sf(poly_path)

ggplot() +
  tidyterra::geom_spatraster_contour_filled(
    data = terra::rast(bathy), 
    breaks = c(seq(0,-600,-100),-10000),
    alpha = 0.6, 
    linewidth = 0.1,
    color = "gray20", show.legend = F) +
  scale_fill_brewer(
    palette = "Blues",
    na.value = "#08519C")  +
  geom_sf(data = new_england) + 
  geom_sf(data = canada) + 
  geom_sf(data = region_extent, fill = "transparent", color = "black", linewidth = 1.2) +
  theme_bw() + 
  map_theme(
    text = element_text(family = "Avenir"),
    plot.title = element_text(face = "bold")) +
  coord_sf(xlim = c(-74.5, -49.5), ylim = c(37, 48)) + # full zoom out to TGB
  labs(title = "Gulf of Maine Study Area")
```



:::


Beginning around 2010, SSTs rose rapidly in the Gulf of Maine and surrounding areas. Annual SSTs
continued to fluctuate, but mean temperatures were hotter than previous decades.[@cr-warmingplot]{scale-by="1.25"} 



:::{#cr-warmingplot}

```{r}

# Timeseries Path
timeseries_path <- region_paths[["apershing_gulf_of_maine"]][["timeseries_path"]]

# Load the regular timeseries
region_timeseries <- read_csv(
  file = timeseries_path, 
  col_types = cols(), 
  guess_max = 1e6) %>% 
  mutate(year = lubridate::year(time))

# Summarize by year to return mean annual anomalies and variance
annual_summary <- region_timeseries %>% 
  filter(year >1981) %>% 
  group_by(year) %>% 
  summarise(across(.cols = -c(time), .fns = mean)) %>% 
  mutate(
    yr_as_dtime = as.Date(paste0(year, "-07-02")),
    anom_direction = ifelse(sst_anom > 0, "Above Long-Term Average", "Below Long-Term Average"))

#RColorBrewer::brewer.pal(9, "RdBu")
# "#B2182B" "#D6604D" "#F4A582" "#FDDBC7" "#F7F7F7" "#D1E5F0" "#92C5DE" "#4393C3" "#2166AC"


ggplot(annual_summary, aes(year, area_wtd_anom)) +
  geom_hline(yintercept = 0, color = "black", alpha = 0.9) +
  geom_col(aes(fill = anom_direction)) +
  geom_smooth(method = "lm", se = F, color = "black") +
  geom_textbox(
    data = data.frame(year = 2017, area_wtd_anom = 0.4),
    aes(label = "Period of Consistently High SSTs After 2010."), 
    family = "Avenir") +
  scale_fill_manual(values = c("#D6604D", "#4393C3")) +
  theme(legend.position = "bottom") +
  scale_y_continuous(labels = label_number(suffix = deg_c)) +
  labs(
    x = "Year", 
    y = "Sea Surface Temperature Anomaly",
    title = "Rapidly Warming Gulf of Maine",
    fill = "")
```

:::


A review of deep-water (250m) temperatures show that temperatures below the surface made a similar transition around the same time.[@cr-buoyplot]{scale-by="1.25"}



:::{#cr-buoyplot}

```{r}
# Package to interface with erddap
library(rerddap)

# Read the data we downloaded from erddap
buoy_m <- read_csv("~/Documents/Repositories/oisst_mainstays/local_data/buoy_m_250m.csv")


# Monthly averages and climatology
bm_month_clim <-  buoy_m  %>% 
  group_by(station, depth, mnth) %>% 
  summarise(
    mean_temp = mean(temperature, na.rm = T),
    .groups = "drop")

# Get Anomalies
bm_monthly <- buoy_m  %>% 
  group_by(station, depth, yr, mnth) %>% 
  summarise(
    temperature = mean(temperature, na.rm = T),
    .groups = "drop") %>% 
  mutate(
    time  = as.Date(
      str_c(yr, str_pad(mnth, pad = "0", side = "left", width = 2), 
            "01", sep = "-"))) %>% 
  left_join(bm_month_clim) %>% 
  mutate(temp_anom = temperature - mean_temp,
         is_pos = if_else(temp_anom < 0, F, T))

# Add the monthly Gaps in explicitly
buoy_month_range <- c(min(bm_monthly$time, na.rm = T), max(bm_monthly$time, na.rm = T))
all_buoy_months <- data.frame(
  "time" = seq.Date(
    from = buoy_month_range[1], 
    to = buoy_month_range[2], 
    by = "month"))

# Fill Gaps
bm_monthly <- all_buoy_months %>%
  left_join(bm_monthly) %>% 
  mutate(anom_col = if_else(is_pos, "#B2182B", "#2166AC"))
# Plot

bm_monthly %>% 
  ggplot() +
  geom_line(aes(time, 0, group = 1)) +
  geom_line(aes(time, temp_anom, color = I(anom_col), group = 1)) +
  geom_segment(
    aes(time, 0, xend = time, yend = temp_anom, group = 1, color = I(anom_col)), alpha = 0.3) +
  geom_line(aes(time, temp_anom, color = I(anom_col), group = 1)) +
  scale_y_continuous(labels = label_number(suffix = deg_c)) +
  labs(
    y = "Temperature Anomaly",
    x = "Date",
    title = "Deep-Water Temperature",
    subtitle = "NERACOOS Buoy M: 250m Temperature Sensor")

```

:::


At the time of Andy's publication, the rapid change in Gulf of Maine SST was attributed in part to a change in the Gulf Stream Position and the Pacific Decadal Oscillation:[@cr-areamap]{scale-by="1.25" pan-to="-30%,15%"}

> The warming, which was related to a northward shift in the Gulf Stream and to changes in the Atlantic Multidecadal Oscillation and Pacific Decadal Oscillation, led to reduced recruitment and increased mortality in the region’s Atlantic cod (Gadus morhua) stock. - Pershing et al. 2015 

Subsequent work has suggested that much of the regional warming may be attributed to natural climate variability:

> ...Finally, the similarity between the basin‐wide average and the shelf/slope meridional mean trends implies approximately two‐thirds of the linear warming trend in 1982–2018 may be
attributed to natural climate variability and approximately one third to external forcing. - Chen et al. 2020

Observations of a Northward shift in the Gulf Stream path have been linked to ocean and atmospheric conditions across the shelf and northward to the Labrador Sea and Greenland (Joyce 2019). Other researchers have noted that in 2008, the Gulf Stream migrated closer to the Tail of the Grand Banks, reducing the westward connectivity of the Labrador Currnet.(Neto 2021) 

:::


### Is it worth it?

Maybe if...

1. There is a defined story you want to tell
2. You are interested in html/css formatting for tweaking the style/layout
3. You are looking to learn more quarto javascript functionality

